---
title: "Disaster Prevention/Addressing Policy: An Analysis of the Most Harmful Natural Events in the US"
author: "Tomas Gogorza"
date: "Wednesday, August 19, 2015"
output: html_document
---

In the last 60 year there's been a countless number of natural disasters which have affected the lives of American citizens in greater or lesser degree. This analysis will classify events by types and show which ones are the most dangerous. Natural disasters may affect the country in several ways, and it is honsidered herby that population health and economy are 2 of the most important factors to take into account when preparing for an event, so this analysis will take main focus on those 2 factors.

#Data Source

The information used on this report was obtained from U.S. National Oceanic and Atmospheric Administration's storm database

```{r cache=TRUE}

setInternet2(use = TRUE)
#Download data from NOAA
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "StormData.csv.bz2")
#Load data
storms <- read.table(bzfile("StormData.csv.bz2"), header = TRUE, sep = ",", na.strings = "", check.names = TRUE, blank.lines.skip = TRUE)  


```

#Data Processing

Libraries used
```{r}
library(dplyr)
```

Clean column names 
```{r cache=TRUE}

names(storms) <- make.names(names(storms), allow_ = FALSE)
names(storms)[1] <- gsub(".","",names(storms)[1],fixed=TRUE)

```

Format dates
```{r}
```

Calculate fatalities by event type
```{r}

fatals <- tapply(storms, FUN = sum, X = FATALITIES, INDEX = storms$EVTYPE)
fatals <- aggregate(storms, FUN = sum, by = list(storms$EVTYPE), x = storms$FATALITIES)

```

Calculate economic damage by event type
```{r}
```


#Results

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?


