---
title: "Tooth Growth Analysis"
author: "Tomas Gogorza"
date: "Tuesday, September 15, 2015"
output: pdf_document
---
#Summary
This document shows a quick analysis and statistical comparison of 2 different supplements in 3 different doses for tooth growth.

#Data Analysis
Load data
```{r warning=FALSE, message=FALSE, error=FALSE}
library(ggplot2)
library(dplyr)

data("ToothGrowth")
data <- as.data.frame(ToothGrowth)
summary(data)
```
From a quick overview of the data, we can see that there are 2 different supplements (OJ & VC) and the dosage range from 0.5 to 2.  
We can plot the samples distribution to get a graphical sense of the difference between supplements and dosage.
```{r}
ggplot(data, aes(x=factor(dose),y=len)) + geom_boxplot() + facet_grid(.~supp) + 
    labs(title="Dose vs Length by Supplement Type", y="Length", x="Dose")
```
**Figure 1.** Distribution of tooth length by dosage, grouped by supplement type  


#Comparison
In order to determine which kind of supplement type is more efficient (if any), we'll perform a dose by dose comparison between them.

##Small Dose (0.5)


##Medium Dose (1)

##Large Dose (2)

#Conclusions and Assumptions