---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Load packages
```{r warning=FALSE, message=FALSE, error=FALSE}

library(ggplot2)
library(dplyr)

```

Load data and clean data set
```{r}

activity <- read.csv("activity.csv")
#activity.withNA <- activity
#activity <- filter(activity,!is.na(steps))

```


## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day:
```{r}

activity.bydate <- aggregate(FUN = sum, by = activity["date"], x=activity["steps"])
activity.bydate


```


2. Make a histogram of the total number of steps taken each day
```{r fig.width=12}

ggplot(na.omit(activity), aes(date,steps)) +
    geom_histogram(stat = "identity") +
    theme(axis.text.x = element_text(angle = 90, vjust=0.5))

```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}

stepstats <- summary(activity.bydate$steps)
stepstats

```

The **mean** total number of steps taken per day is **`r as.integer(stepstats['Mean'])`** steps/day and the **median** is **`r as.integer(stepstats['Median'])`** steps/day


## What is the average daily activity pattern?

Average data by daily time interval
```{r}

activity.byinterval <- aggregate(FUN = mean, by = activity["interval"], x=activity["steps"], na.rm=TRUE)

```

1. Make time series plot (avg steps taken per interval, across all days)
```{r}

ggplot(na.omit(activity.byinterval), aes(interval,steps)) + 
  geom_line() + 
  scale_x_continuous(breaks = seq(min(activity.byinterval$interval), 
                                  max(activity.byinterval$interval), by = 200))

```

2. Get the interval with max average steps
```{r}

maxsteps <- summary(activity.byinterval$steps)["Max."]
maxinterval <- filter(activity.byinterval, as.integer(steps)==as.integer(maxsteps))

```

The time interval averaging the maximum number of steps is the **`r maxinterval$interval`** interval


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset
```{r}



```

## Are there differences in activity patterns between weekdays and weekends?
